Ace: An Actively-Typed Language and Compilation Environment for High-Performance Computing

### 2013-04-26 ace/ace.tex
Revised up through type propagation

\subsection{Phase Separation}
\begin{codelisting}
\lstinputlisting{listing3.py}
\caption{[\texttt{listing\ref{map}.py}] A generic data-parallel higher-order map function written using the OpenCL user module.}
\label{map}
\end{codelisting}
\begin{codelisting}
\lstinputlisting{listing4.py}
\caption{[\texttt{listing\ref{mapadd5dbl}.py}] The generic \texttt{map} function compiled to map the \texttt{add5} function over two  types of input.}
\label{mapadd5dbl}
\end{codelisting}
\begin{codelisting}
\lstinputlisting[style=OpenCL]{listing5.cl}
\caption{[\texttt{listing\ref{mapout}.cl}] The OpenCL code generated by running \texttt{acec listing2.py}.}
\label{mapout}
\end{codelisting}

@@

\subsection{Example 2: Higher-Order Map}
The ``Hello, World!'' example demonstrates the structure of Ace programs, but it does not require working with types. Listing \ref{map} shows an imperative, data-parallel map primitive written using the OpenCL library introduced above. To review, in OpenCL users can define functions, called {\em kernels}, that execute across thousands of threads. Each kernel has access to a unique index, called its {\em global id}, which can be used by the programmer to ensure that each thread operates on different parts of the input data (Line 5). The \verb|map| kernel defined in Listing \ref{map} applies a transfer function, \verb|f|, to the element of the input array, \verb|input|, corresponding to its global id. It writes the result of this call into the corresponding location in the provided output array, \verb|output|.

As above, \verb|map| is a {\em generic function} (specifically, it is an instance of the class \verb|ace.GenericFn|). This means that it's arguments have not been assigned types. Indeed, the functionality given by the \verb|map| definition is applicable to many combinations of types for \verb|input| and \verb|output| and functions, \verb|f|. In this sense, \verb|map| is actually a family of functions defined for all types assignments for \verb|input|, \verb|output| and \verb|f| for which the operations in the function's body are well-defined.

Running \verb|acec listing3.py| would produce no output, however. To create a {\em concrete function} (an instance of the class \verb|ace.ConcreteFn|) that can be emitted by the compiler, types must be assigned to each of the arguments of any externally callable functions. Listing \ref{mapadd5dbl} shows how to use the \verb|compile| method to specialize \verb|map| in two different ways to apply the \verb|add5| function, defined on Lines 4-6, to arrays  that reside in global memory (OpenCL has a notion of four different memory spaces). Line 9 produces a version specialized for arrays of \verb|double|s and Line 10 produces a version for arrays of \verb|int|s. The output of compilation is shown in Figure \ref{mapout}.

\subsection{Types as Metalanguage Objects}
The types of the arguments of the functions being compiled on Lines 4.9 and 4.10, if written directly in OpenCL as in Listing \ref{mapout}, are \verb|__global double*| and \verb|__global int*|, respectively. The corresponding types in Ace are \verb|global_ptr(double)| and \verb|global_ptr(int)|, abbreviated for convenience as \verb|A| and \verb|B| respectively on Line 8 (note that here, standard variable assignment in the metalanguage is serving the role that a \verb|typedef| declaration would fill in a C-like language.)

\textbf{Types in Ace are values in the metalanguage}, Python. More specifically, types are {\em object instances} of user-defined classes that inherit from the Ace-provided \verb|ace.Type| class. For example, \verb|global_ptr(double)| is an instance of \verb|ace.OpenCL.GlobalPtrType| instantiated with the target type, \verb|double|, as a constructor argument. The types \verb|double| and \verb|int| (imported from \verb|ace.OpenCL| on Line 4.2) are instances of \verb|ace.OpenCL.FloatType| and \verb|ace.OpenCL.IntegerType|, respectively. As we will describe further in Section \ref{att}, this notion of types as metalanguage objects is a key to Ace's flexibility.

\subsection{Type Propagation and Higher-Order Functions}
The type assigned to the third argument, \verb|f|, on Lines 4.9 and 4.10, is given as \verb|add5.ace_type|. The \verb|ace_type| attribute of a generic function is an instance of \verb|ace.GenericFnType|, the type of Ace generic functions, that corresponds to that particular function, \verb|add5| in this case. Generic functions are compiled automatically when they are called from another function. That is, when the compiler encounters the call to \verb|f| inside \verb|map| when compiling \verb|map_add5_double|, it compiles a version of \verb|add5| specialized to the \verb|double| type (seen on Line 5.3), and similarly with the \verb|int| type when compiling \verb|map_add5_int| (on Line 5.16, automatically given a unique name to avoid conflicts). This mechanism is called {\em type propagation}. In other words, we did not need to use \verb|add5.compile(double)| before compiling \verb|map_add5_dbl| (although we could if we would like). Only functions that are never called in the process of compiling other functions in a module need type information explicitly provided, whereas \verb|add5| is a function that is only called within \verb|map| in Listing \ref{mapadd5dbl}.

We note that this scheme allows for a form of higher-order functional programming even when targeting languages, like OpenCL, that have no support for higher-order functions (OpenCL, unlike C99, does not even support function pointers). This works because the \verb|ace.GenericFnType| for one function, such as \verb|add5|, is not the same as the \verb|ace.GenericFnType| for a superficially similar function, such as \verb|add6| (defined as one would expect). To put it in type theoretic terms, \verb|ace.GenericFnType|s are singleton types, {\em uniquely inhabited} by a single generic function, and thus it is impractical to use them as full first-class values (i.e. they cannot put into a run-time array.) In practice, this is rarely a problem, particularly in parallel programming where explicit specialization is already common in order to avoid the potential performance penalty associated with a function pointer dereference per call.
We note that fully-featured first-class functions can be implemented when targeting a backend that supports them, such as C99, or simulated by using a jump table implementation in OpenCL, but we do not discuss this further due to lack of space.

@@

\begin{codelisting}
\lstinputlisting{listing6.py}
\caption{\texttt{[listing6.py]} A function demonstrating whole-function type inference when multiple values with differing types are assigned to a single variable.}
\label{inference}
\end{codelisting}

>>>>

\subsection{Phase Separation}
\begin{codelisting}
\lstinputlisting{listing3.py}
\caption{[\texttt{listing\ref{map}.py}] A generic data-parallel higher-order map function written using the OpenCL user module.}
\label{map}
\end{codelisting}
\begin{codelisting}
\lstinputlisting{listing4.py}
\caption{[\texttt{listing\ref{mapadd5dbl}.py}] The generic \texttt{map} function compiled to map the \texttt{add5} function over two  types of input.}
\label{mapadd5dbl}
\end{codelisting}
\begin{codelisting}
\lstinputlisting[style=OpenCL]{listing5.cl}
\caption{[\texttt{listing\ref{mapadd5dbl}.cl}] The OpenCL code generated by running \texttt{acec listing\ref{mapadd5dbl}.py}.}
\label{mapout}
\end{codelisting}

@@

\subsection{Example 2: Higher-Order Map}
The ``Hello, World!'' example demonstrates the structure of Ace programs, but it does not require working with types. Listing \ref{map} shows an imperative, data-parallel map primitive written using the OpenCL library introduced above. To review, in OpenCL users can define functions, called {\em kernels}, that execute across thousands of threads. Each kernel has access to a unique index, called its {\em global id}, which can be to ensure that each thread operates on different parts of the input data (Line 5). The \verb|map| kernel defined in Listing \ref{map} applies a transfer function, \verb|f|, to the element of the input array, \verb|input|, corresponding to its global id. It writes the result of this call into the corresponding location in the provided output array, \verb|output|.

As above, \verb|map| is a {\em generic function} (specifically, it is an instance of the class \verb|ace.GenericFn|). This means that its arguments have not been assigned types. The functionality given by the \verb|map| definition is in fact applicable to many combinations of types for \verb|input| and \verb|output| and functions, \verb|f|. In this sense, \verb|map| is actually a \emph{family} of functions defined for all types assignments for \verb|input|, \verb|output| and \verb|f| such that the operations in the function's body are well-defined.

Running \verb|acec listing3.py| would produce no output. To create a {\em concrete function} (that is, an instance of the class \verb|ace.ConcreteFn|) that can be emitted by the compiler, types must be assigned to each of the arguments. Listing \ref{mapadd5dbl} shows how to use the \verb|compile| method to specialize \verb|map| in \emph{two} different ways to apply the \verb|add5| function, defined on Lines 4-6, to arrays  that reside in global memory (OpenCL associates a memory space with pointer types). Line 9 results in a version specialized for arrays of \verb|double|s and Line 10 results in a version for arrays of \verb|int|s. The output of compilation is shown in Listing \ref{mapout}.

\subsection{Types as Metalanguage Objects}
The \verb|compile| method assigns types to the arguments of a generic function. In Listing \ref{mapadd5dbl}, the types we are using are given shoter names, for convenience, on Line 8 (that is, variables in the metalanguage can be used like \verb|typedef| is used in a C-like language). The types \verb|int| and \verb|double| imported from the \verb|OpenCL| module correspond to the OpenCL types of the same name. The types \verb|gptr(int)| and \verb|gptr(double)| correspond to \verb|__global int*| and \verb|__global double*|. That is, \verb|gptr| can be understood as a type-indexed family of types.

These types are \emph{objects in the metalanguage}, Python. More specifically, types are {instances} of user-defined classes that inherit from the Ace-provided \verb|ace.Type| class. For example, \verb|gptr(double)| is an instance of \verb|OpenCL.GlobalPtrType| instantiated with the target type, \verb|double|, as a constructor argument. The types \verb|double| and \verb|int| are instances of \verb|OpenCL.FloatType| and \verb|OpenCL.IntegerType|, respectively. This notion of types as metalanguage objects is key to the Ace compilation model and also enables other mechanisms that we will discuss in subsequent sections.

\subsection{Type Propagation}
\begin{codelisting}
\lstinputlisting{listing6.py}
\caption{\texttt{[listing6.py]} A function demonstrating whole-function type inference when multiple values with differing types are assigned to a single variable.}
\label{inference}
\end{codelisting}
The type assigned to the third argument, \verb|f|, on both Lines 4.9 and 4.10, is \verb|add5.ace_type|. The \verb|ace_type| attribute of a generic function is an instance of \verb|ace.GenericFnType|, the type of Ace generic functions. Ace generic functions are compiled to concrete functions automatically at all internal call sites. That is, when the compiler encounters the call to \verb|f| inside \verb|map| when compiling \verb|map_add5_double|, it compiles a version of \verb|add5| specialized to the \verb|double| type (seen on Line 5.3), and similarly when compiling \verb|map_add5_int| (on Line 5.16, automatically given a unique name to avoid conflicts). This mechanism is called {\em type propagation}. We did not need to use \verb|add5.compile(double)| before compiling \verb|map_add5_dbl| because only functions that are never called in the process of compiling other functions in a module need type information explicitly provided, supporting \textbf{ease-of-use} by increasing conciseness.

In effect, this scheme allows for a form of higher-order functional programming even when targeting languages, like OpenCL, that have no support for higher-order functions (OpenCL, unlike C99, does not support function pointers). This works because the \verb|ace.GenericFnType| for one function, such as \verb|add5|, is not equal to  the \verb|ace.GenericFnType| for a superficially similar function, such as \verb|add6| (defined as one would expect). To put it in type theoretic terms, \verb|ace.GenericFnType|s are singleton types, {\em uniquely inhabited} by a single generic function. A consequence of this is that they cannot be used as first-class values (i.e. they cannot written into an array). This is often valuable, particularly in parallel programming where compile-time specialization is valuable to avoid the \textbf{performance} and \textbf{ease-of-use} issues that occur if using function pointers.
Concrete functions, on the other hand, can be given a true function type (e.g. \verb|add5| could be compiled to a concrete function with type \verb|int| $\rightarrow$ \verb|int|) if targeting a backend that supports them, such as C99, or by using an integer-indexed jump table in OpenCL (we have not implemented this mechanism using Ace as of the time of writing, but do not anticipate difficulties).

Type propagation via generic functions can be compared to template specialization in C++, where both the necessary template headers (containing nested template parameters to support function passing) and template specializations are inferred automatically from usage. This significantly simplifies a sophisticated feature of C++ and introduces it to OpenCL and C, which do not support templates.
% Is this really any more concise than using C + templates? Etc. relation to types. Conciseness.

@@
